---
title: "3. Interactive Dashboards with Shinylive I"
format:
  html:
    other-links:
      - text: Lecture Slides
        icon: file-slides
        href: ../slides/w9-slides-lecture.html
    format-links: false
---

Last week, we used `hoopR` to source NBA data and save reusable files. This week we move from **data sourcing** into the next part of the workflow: **visualisation and a first interactive dashboard**.

The worked example is a **Lakers Evolution Dashboard**. On 1 February 2025 the Los Angeles Lakers traded for Luka Dončić. We will use that mid-season change as a natural before/after split and ask a simple, focused question: *did each Lakers role player's shot selection — the mix of zones they shoot from — change between the AD Era and the Luka Era?*

We will:

1.  read a prepared shot file with several Lakers role players across four seasons;
2.  split those seasons into an **AD Era** (before the trade) and a **Luka Era** (from the trade onwards);
3.  build a clean **static bar chart** comparing shot-zone shares across the two eras;
4.  convert the static chart into an **interactive Plotly** chart with hover tooltips;
5.  **pre-aggregate** the data to only what the chart needs, so the Shiny app stays lightweight;
6.  turn the chart into a **Shiny app**, built up one feature at a time, where the user picks the player.

## Recap: From API Work to Dashboard Work

Last week's workflow was:

``` text
documentation -> pull -> inspect -> clean -> save
```

This week we use the file created by that workflow. Rather than asking everyone to re-run NBA API calls in class, we will start from a prepared `.rds` file. That lets us spend the seminar time on charts, interactivity and app structure.

The angle for this example is:

> **Shot-diet lens:** the Luka trade reshaped who the Lakers run their offence through. For each role player, did the *mix of zones* they shoot from change between the AD Era and the Luka Era?

`shot_zone_basic` is the right column for that question. It is a tidy NBA-Stats label that groups every shot into a familiar court area — *Restricted Area*, *In The Paint (Non-RA)*, *Mid-Range*, *Left Corner 3*, *Right Corner 3*, *Above the Break 3*. If a role player's shot diet shifts across those zones after the trade, a simple bar chart will show it.

The Lakers are a useful case study because the trade gives us a natural before/after with the *same players* on either side. We will look at five role players who spent meaningful time on the Lakers roster around the trade:

-   Rui Hachimura
-   Austin Reaves
-   LeBron James
-   Jaxson Hayes
-   Dorian Finney-Smith

To get a more reliable picture of each player's shot diet — rather than relying on a half-season on either side of the trade — the prepared file pulls **four NBA regular seasons**: 2022/23, 2023/24, 2024/25 and 2025/26. Each player's *AD Era* window therefore includes the Lakers seasons they played **before** 1 February 2025, and the *Luka Era* includes the rest of 2024/25 plus all of 2025/26. If a player wasn't on the Lakers for a particular season, that season is simply absent from their data.

The dashboard question is deliberately focused. We are not trying to explain the whole NBA. We are building a small interactive tool that shows how each Lakers role player's shot selection shifted after Luka arrived.

## From Static Charts to a Shiny App

By the end of this workbook, you will have:

-   read a prepared `.rds` file with shots from five Lakers role players across four seasons;
-   added an `era` column based on the Luka trade date;
-   built a static bar chart comparing each player's shot distribution across the two eras;
-   converted the static chart into a Plotly chart with hover text;
-   saved a small pre-aggregated dataset for the app to read;
-   built a first Shiny app where a dropdown picks the player;
-   added a sidebar layout, cleaner titles and Plotly *inside* the Shiny app.

The important sequence is:

``` text
readRDS() -> derive era -> static bar plot -> Plotly bar plot -> pre-aggregate -> Shiny app
```

Each step adds one idea. If a later step breaks, the earlier step still works — which makes debugging much easier.

### Before We Start: Check the Data Folders

Before downloading today's cleaned Lakers file, tidy the `data/` folder you created last week. Downloaded API data can become large and change when you re-run calls, so it belongs in `data/raw/`. Cleaned tables that you use for charts and apps belong in `data/cleaned/`.

Start in the **Files** pane in RStudio. It is usually in the bottom-right panel.

1.  Find your project files in the **Files** pane.
2.  Click the `data` folder to open it.
3.  Click `New Folder`.
4.  Type `raw`.
5.  Click `OK` or `Create`.
6.  Create a second folder called `cleaned`.

You should now have:

``` text
data/
  raw/
  cleaned/
```

If you prefer to create the folder with code, this does the same thing:

```{r}
#| eval: false
#| filename: "R code"

dir.create("data/raw", recursive = TRUE, showWarnings = FALSE)
dir.create("data/cleaned", recursive = TRUE, showWarnings = FALSE)
```

Next, move any existing files from the top level of `data/` into the right subfolder.

In the **Files** pane:

1.  Stay inside the `data` folder.
2.  Tick downloaded API files, such as `nba_2024_team_box.rds`, and move them into `data/raw/`.
3.  Click `More`.
4.  Choose `Move`.
5.  Select the target folder.
6.  Click `Move`.

Cleaned summary files, such as `nba_2024_team_stats.rds`, should move into `data/cleaned/`.

You can also move the existing files with code:

```{r}
#| eval: false
#| filename: "R code"

items_to_move <- list.files("data", full.names = TRUE, recursive = FALSE)
items_to_move <- items_to_move[file.info(items_to_move)$isdir == FALSE]

raw_files <- items_to_move[grepl("_box\\.(rds|csv)$", basename(items_to_move))]
cleaned_files <- items_to_move[grepl("_stats\\.(rds|csv)$", basename(items_to_move))]

if (length(raw_files) > 0) {
  file.rename(raw_files, file.path("data/raw", basename(raw_files)))
}

if (length(cleaned_files) > 0) {
  file.rename(cleaned_files, file.path("data/cleaned", basename(cleaned_files)))
}
```

Now open `.gitignore` and add this line:

``` text
data/raw/
```

This means Git will ignore the raw downloaded files. That is good practice because raw sports data can be large, slow to regenerate and unnecessary to store in every commit. We will work from a cleaned `.rds` file instead.

### Get the Prepared Data

Download the cleaned Lakers shot file from [Canvas](https://canvas.liverpool.ac.uk/courses/85669/files/13865218?module_item_id=2580816) and place it in your project folder like this:

``` text
data/
  cleaned/
    lakers_shots.rds
```

The file was created using the same technique we covered in [Workbook 2](02-api.html): read the documentation, choose the right parameters, pull the data, inspect the returned columns, clean the table, and save the result. Each row is one field-goal attempt by one of the five Lakers role players during the selected regular seasons.

::: callout-tip
If you are interested, open the optional section below in your own time. You do not need to run those API calls during the seminar. For today, the practical task starts from the prepared `.rds` file.
:::

::: {.callout-note collapse="true"}
### Optional: how this RDS file was created

The code below shows the data-cleaning route used to create the Canvas file. This is here for transparency, not because you need to repeat it in class.

**Packages and constants**

```{r}
#| eval: false
#| filename: "R code"

library(hoopR)
library(dplyr)
library(purrr)
library(tidyr)

lakers_players <- tibble::tribble(
  ~player_id, ~player_name,
  1629060,    "Rui Hachimura",
  1630559,    "Austin Reaves",
  2544,       "LeBron James",
  1629637,    "Jaxson Hayes",
  1627827,    "Dorian Finney-Smith"
)

nba_stats_seasons <- c("2022-23", "2023-24", "2024-25", "2025-26")
lakers_team_id    <- 1610612747
```

Four seasons gives us a more reliable sample on either side of the trade. The `lakers_team_id` filter (NBA-Stats' team ID for the Lakers) is passed through to the API call so that, for a player who was traded in and out of Los Angeles, we only keep the shots they took *in a Lakers jersey*.

**Small helper functions**

```{r}
#| eval: false
#| filename: "R code"

parse_nba_date <- function(x) {
  x_chr <- as.character(x)
  parsed <- rep(as.Date(NA), length(x_chr))
  nba_date <- grepl("^\\d{8}$", x_chr)

  parsed[nba_date]  <- as.Date(x_chr[nba_date], format = "%Y%m%d")
  parsed[!nba_date] <- suppressWarnings(as.Date(x_chr[!nba_date]))

  parsed
}

get_player_shots <- function(player_id, player_name, season_label) {
  res <- tryCatch(
    hoopR::nba_shotchartdetail(
      player_id       = player_id,
      season          = season_label,
      season_type     = "Regular Season",
      team_id         = lakers_team_id,
      context_measure = "FGA"
    ),
    error = function(e) NULL
  )

  if (is.null(res) || is.null(res$Shot_Chart_Detail) ||
      nrow(res$Shot_Chart_Detail) == 0) {
    return(NULL)
  }

  res$Shot_Chart_Detail |>
    dplyr::mutate(season_label = season_label)
}
```

`nba_shotchartdetail()` returns one player's shot chart at a time, so we wrap it in a small helper and call it once per player-season combination. The `tryCatch(...)` plus the `NULL` return handles the player-seasons where a player wasn't on the Lakers — the API just gives us back an empty table and we skip it.

**Loop over players and bind**

```{r}
#| eval: false
#| filename: "R code"

shot_grid <- tidyr::expand_grid(
  lakers_players,
  season_label = nba_stats_seasons
)

lakers_shots_raw <- purrr::pmap(shot_grid, get_player_shots) |>
  purrr::compact() |>
  dplyr::bind_rows()

dir.create("data/raw", recursive = TRUE, showWarnings = FALSE)
saveRDS(lakers_shots_raw, "data/raw/lakers_shots_raw.rds")
```

`expand_grid()` produces every player × season combination. `pmap()` calls `get_player_shots()` on each row, `compact()` drops the empty results (for player-seasons that weren't on the Lakers), and `bind_rows()` stacks what's left.

**Clean and save**

```{r}
#| eval: false
#| filename: "R code"

luka_trade_date <- as.Date("2025-02-01")

lakers_shots <- lakers_shots_raw |>
  rename_with(tolower) |>
  transmute(
    player_id,
    player_name,
    game_id,
    game_date    = parse_nba_date(game_date),
    season = as.integer(substr(season_label, 1, 
  4)) + 1L,
    season_label,
    x            = as.numeric(loc_x),
    y            = as.numeric(loc_y),
    shot_type,
    action_type,
    shot_zone_basic,
    shot_value     = if_else(shot_type == "3PT Field Goal", 3L, 2L),
    shot_made_flag = as.integer(shot_made_flag),
    era = if_else(game_date < luka_trade_date, "AD Era", "Luka Era")
  ) |>
  mutate(
    era = factor(era, levels = c("AD Era", "Luka Era"))
  )

dir.create("data/cleaned", recursive = TRUE, showWarnings = FALSE)
saveRDS(lakers_shots, "data/cleaned/lakers_shots.rds")
```

The cleaned table has one row per shot. The `era` column is the key new variable: it splits each player's season into the AD Era (before 1 February 2025) and the Luka Era (from 1 February 2025 onwards). That single column is what makes the dashboard's before/after comparison possible.
:::

### Load the Data for Today

We only need a small set of packages for today's chart and app.

```{r}
#| message: false
#| warning: false
#| filename: "R code"

library(dplyr)
library(ggplot2)
library(plotly)
library(shiny)
```

Read the prepared file, make sure `era` is in place, and put `shot_zone_basic` in a sensible left-to-right order (close range → long range).

```{r}
#| filename: "R code"

luka_trade_date <- as.Date("2025-02-01")

zone_order <- c(
  "Restricted Area",
  "In The Paint (Non-RA)",
  "Mid-Range",
  "Left Corner 3",
  "Right Corner 3",
  "Above the Break 3"
)

lakers_shots <- readRDS("data/cleaned/lakers_shots.rds") |>
  mutate(
    game_date = as.Date(game_date),
    era = factor(
      if_else(game_date < luka_trade_date, "AD Era", "Luka Era"),
      levels = c("AD Era", "Luka Era")
    )
  ) |>
  filter(shot_zone_basic %in% zone_order) |>
  mutate(shot_zone_basic = factor(shot_zone_basic, levels = zone_order))
```

Two small choices are doing useful work here:

-   `era` is a **factor** with explicit level order so the AD Era always appears to the left of the Luka Era in every chart and legend.
-   `shot_zone_basic` is also a factor with `zone_order`. That means bars read naturally from close-range on the left to long-range on the right, no matter which player we plot.

Check the shape of the data before plotting.

```{r}
#| filename: "R code"

glimpse(lakers_shots)
count(lakers_shots, player_name, era)
```

Each row is one field-goal attempt. The second table confirms every player has shots in **both** eras — the condition for a meaningful before/after view.

### A Static Bar Chart with ggplot2

Before making anything interactive, build the chart as a normal static figure. A static figure is easy to inspect, easy to debug and easy to share in a report.

We will start with one player — Rui Hachimura — and one question: *what share of his shots came from each zone, in each era?*

**Step 1:** compute the per-era share of shots in each zone.

```{r}
#| filename: "R code"

rui_by_zone <- lakers_shots |>
  filter(player_name == "Rui Hachimura") |>
  count(era, shot_zone_basic) |>
  group_by(era) |>
  mutate(share = n / sum(n)) |>
  ungroup()

rui_by_zone
```

For each era we count shots per zone, then divide by the era total. The result is a tidy table with one row per (era, zone) pair and a `share` column ready for plotting.

**Step 2:** turn that table into a clean grouped bar chart.

```{r}
#| label: fig-rui-static-bars
#| fig-cap: "Rui Hachimura: share of shot attempts by zone, AD Era vs Luka Era."
#| filename: "R code"

lakers_colours <- c("AD Era" = "#FDB927", "Luka Era" = "#552583")

ggplot(rui_by_zone,
       aes(x = shot_zone_basic, y = share, fill = era)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = lakers_colours) +
  labs(
    title    = "Rui Hachimura: shot distribution by zone",
    subtitle = "Share of attempts from each zone, AD Era vs Luka Era",
    x = NULL, y = "Share of attempts", fill = NULL,
    caption  = "Data: NBA Stats. AD Era = before 1 Feb 2025, Luka Era = from 1 Feb 2025."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "top",
    plot.title         = element_text(face = "bold"),
    axis.text.x        = element_text(angle = 20, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

A few small design choices make the plot read cleanly:

-   **Lakers colours.** Gold for the AD Era, purple for the Luka Era — instantly readable to a Lakers audience.
-   **Dodged bars.** `position_dodge()` puts the two eras side by side within each zone, which is the comparison we want.
-   **Percentage y-axis.** `scales::percent_format()` prints `25%` rather than `0.25`.
-   **Minimal theme.** `theme_minimal()` plus a few tweaks (no vertical gridlines, angled zone labels) keeps the chart tidy without fighting the data.

The bar chart already tells a story: any zone where the purple bar is taller than the gold one is a zone Rui leans into *more* after the trade, and vice versa. What we cannot yet do is switch players, or read an exact percentage off a bar. That is where Plotly and Shiny come in.

### Make the Figure Interactive with Plotly

Plotly adds three things you do not get from a static chart:

-   **hover text**, so the viewer can read the exact share of each bar;
-   **zoom and pan**, useful when you have lots of categories or values;
-   **legend toggling**, so you can click a legend entry to hide or show a series.

`plotly::ggplotly()` will convert the existing `ggplot` object directly. We add one extra piece inside `aes()` — a `text` aesthetic that describes the tooltip — and pass `tooltip = "text"` to `ggplotly()` so Plotly uses only that field on hover.

```{r}
#| label: fig-rui-plotly-bars
#| filename: "R code"

rui_plot <- ggplot(rui_by_zone,
                   aes(x = shot_zone_basic, y = share, fill = era,
                       text = paste0(
                         era, "<br>",
                         shot_zone_basic, "<br>",
                         "Attempts: ", n, "<br>",
                         "Share: ", round(share * 100, 1), "%"
                       ))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = lakers_colours) +
  labs(fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))

ggplotly(rui_plot, tooltip = "text")
```

Two lines are doing most of the work:

-   `text = paste0(...)` inside `aes()` defines *what the hover shows*. ggplot itself ignores the `text` aesthetic, but `ggplotly()` picks it up.
-   `ggplotly(..., tooltip = "text")` tells Plotly to use only that field for the tooltip, instead of listing every mapped aesthetic.

The chart is now exploratory: a coach can hover any bar to read the exact share and number of attempts. There is still one problem — the chart only knows about Rui. To compare Austin Reaves, we would have to rerun the code by hand. That is exactly what Shiny is for.

### Prepare a Smaller Dataset for Shiny

Before we move to Shiny, a small but important step: **pre-aggregate the data**.

The raw `lakers_shots` table has one row per shot — tens of thousands of rows across four seasons. Our chart does not need any of those rows; it only needs, for each player × era × zone, the **count of attempts** and the **share**. That is 5 players × 2 eras × 6 zones = at most **60 rows** of summary data.

Why does this matter for Shiny?

-   **Loading time.** When a user visits the app, Shiny reads the data file into memory. A 100-KB summary loads instantly; a 10-MB shot file is overkill when no individual shot is ever shown.
-   **Per-click speed.** Every time the user changes the dropdown, the server re-runs. If the aggregation happens inside the app, it runs on every click. If we aggregate ahead of time, the app only has to `filter(player_name == input$player)`.
-   **Deployment.** When you share or deploy the app later, only files inside the `app/` folder travel with it. Smaller files mean a smaller, faster deployment.

The rule of thumb: **do the heavy lifting before the app, and ship only what the chart needs.**

Compute the summary and save it into `app/shot_zones.rds`.

```{r}
#| filename: "R code"

shot_zones <- lakers_shots |>
  count(player_name, era, shot_zone_basic) |>
  group_by(player_name, era) |>
  mutate(share = n / sum(n)) |>
  ungroup()

shot_zones
```

Each row is one combination of player, era and zone, with the attempt count and the share within that era. That is the exact shape the chart consumes.

Create the `app/` folder if it does not already exist, then write `shot_zones.rds` into it.

```{r}
#| eval: false
#| filename: "R code"

dir.create("app", showWarnings = FALSE)
saveRDS(shot_zones, "app/shot_zones.rds")
```

```{r}
#| include: false

dir.create("app", showWarnings = FALSE)
saveRDS(shot_zones, "app/shot_zones.rds")
```

The app will read **this** file — not the original per-shot file. From here on, we can forget about `lakers_shots` entirely.

## Turn It Into a Shiny App

Plotly makes *a chart* interactive. Shiny makes the *whole page* reactive: the user clicks something, and R recomputes the chart for them. Instead of writing one chart per player, you write one chart *template* and let the user choose.

We will build the app in four small versions. Each version adds one feature. You can copy each version into `app/app.R`, run it, then copy the next version over the top.

### What is Shiny?

A Shiny app is just an `app.R` script with two objects:

-   a **UI** (`ui`), which describes the inputs and outputs — what the user sees on the page;
-   a **server** function (`server`), which describes what R should compute whenever the inputs change.

`shiny::runApp("app")` launches the script as a small web app in your browser.

### A Quick Reminder About Working Directories

One quick detour before writing the app. Every R session has a **working directory** — the folder R looks in when you write a relative path like `"shot_zones.rds"`.

-   Opening the `.Rproj` file sets the working directory to the **project root** (where `data/` and `app/` live).
-   `shiny::runApp("app")` temporarily changes the working directory into the `app/` folder for the lifetime of the app.

That is why the app reads `"shot_zones.rds"` with no folder prefix: once the app is running, `app/` *is* the working directory. The pre-aggregation step above already wrote `app/shot_zones.rds`, so the app has what it needs. If you ever rebuild `shot_zones`, rerun the `saveRDS(..., "app/shot_zones.rds")` line to refresh it.

Your project should look like this:

``` text
app/
  shot_zones.rds
data/
  cleaned/
    lakers_shots.rds
  raw/
    lakers_shots_raw.rds
```

The raw API pull stays in `data/raw/` if you generated it yourself. The cleaned per-shot file stays in `data/cleaned/` as the reusable archive. Only the 60-row summary travels with the app.

### Version 1 — A Minimal App

Version 1 is the smallest useful Shiny app: no dropdowns, no title, no layout. Just the bar chart we already built, wrapped in the Shiny page/server machinery.

```{r}
#| eval: false
#| filename: "app/app.R"

library(shiny)
library(dplyr)
library(ggplot2)

lakers_colours <- c("AD Era" = "#FDB927", "Luka Era" = "#552583")

shot_zones <- readRDS("shot_zones.rds")

ui <- fluidPage(
  plotOutput("zone_bars")
)

server <- function(input, output, session) {
  output$zone_bars <- renderPlot({
    shot_zones |>
      filter(player_name == "Rui Hachimura") |>
      ggplot(aes(x = shot_zone_basic, y = share, fill = era)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      scale_fill_manual(values = lakers_colours) +
      labs(title = "Rui Hachimura: shot distribution by zone",
           x = NULL, y = "Share of attempts", fill = NULL) +
      theme_minimal(base_size = 12) +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))
  })
}

shinyApp(ui = ui, server = server)
```

Run the app:

```{r}
#| eval: false
#| filename: "R Console"

shiny::runApp("app")
```

Three things to notice:

-   `plotOutput("zone_bars")` in the UI is a **placeholder**. It says, "put a plot here, and call it `zone_bars`".
-   `output$zone_bars <- renderPlot({ ... })` in the server **fills that placeholder** with whatever the block returns.
-   The player name is still hard-coded. The app is not yet interactive — we are only reusing Shiny's page and plot machinery.

That is Shiny in its simplest form. Everything we add from here on plugs into those same two objects.

### Version 2 — Add a Player Selector

Now the interesting move: let the user choose the player. We add two things:

1.  a `selectInput()` in the UI, called `"player"`;
2.  a reference to `input$player` in the server, where the name used to be hard-coded.

Only the UI block and a couple of lines in the server change. The rest of the script is identical to Version 1.

```{r}
#| eval: false
#| filename: "app/app.R (changes only)"

player_choices <- sort(unique(shot_zones$player_name))

ui <- fluidPage(
  selectInput(
    inputId  = "player",
    label    = "Player",
    choices  = player_choices,
    selected = "Rui Hachimura"
  ),
  plotOutput("zone_bars")
)

server <- function(input, output, session) {
  output$zone_bars <- renderPlot({
    shot_zones |>
      filter(player_name == input$player) |>
      ggplot(aes(x = shot_zone_basic, y = share, fill = era)) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      scale_fill_manual(values = lakers_colours) +
      labs(fill = NULL) +
      theme_minimal(base_size = 12) +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))
  })
}
```

Run the app again. Changing the dropdown now changes the chart — and the title updates to the selected player too. Two ideas underneath this:

-   `input$player` holds the **current value** of the dropdown. Whenever that value changes, Shiny marks any output that reads `input$player` as *stale* and re-runs it.
-   Building `paste0(input$player, ": ...")` into the chart title is a small but important touch. A headline that changes with the selection makes the app feel responsive.

That is reactivity at its simplest: **an output that reads an input automatically re-runs when the input changes**.

### Version 3 — A Cleaner Layout

Version 2 works, but the dropdown sits awkwardly above a huge plot. A typical Shiny layout uses `sidebarLayout()`: inputs in a narrow sidebar on the left, outputs in the main area on the right. We also add a `titlePanel()` at the top and a short sentence in the sidebar to explain the chart.

```{r}
#| eval: false
#| filename: "app/app.R (UI only)"

ui <- fluidPage(
  titlePanel("Lakers Evolution Dashboard"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      selectInput(
        inputId  = "player",
        label    = "Choose a player",
        choices  = player_choices,
        selected = "Rui Hachimura"
      ),
      p("Each bar shows the share of that player's attempts from a given zone.
         Compare the AD Era (before 1 Feb 2025) to the Luka Era (from 1 Feb 2025).")
    ),
    mainPanel(
      width = 9,
      plotOutput("zone_bars", height = "500px")
    )
  )
)
```

The server block is unchanged. All we did was rearrange the page.

This is an important idea. Once reactivity is wired up, **layout work is separate from data work**. You can spend time polishing `sidebarLayout()`, headings, spacing and explanatory text *without* touching the server — and without breaking anything.

### Version 4 — Plotly Inside Shiny

Swapping the static plot for an interactive Plotly chart inside Shiny takes three small changes:

1.  `plotOutput("zone_bars")` becomes `plotlyOutput("zone_bars")`.
2.  `renderPlot({ ... })` becomes `renderPlotly({ ... })`.
3.  Wrap the final `ggplot(...)` in `ggplotly(..., tooltip = "text")`, and add a `text = ...` aesthetic so the tooltip shows the share and number of attempts.

Here is the full `app.R` for Version 4:

```{r}
#| eval: false
#| filename: "app/app.R"

library(shiny)
library(plotly)
library(dplyr)
library(ggplot2)

lakers_colours <- c("AD Era" = "#FDB927", "Luka Era" = "#552583")

shot_zones <- readRDS("shot_zones.rds")

player_choices <- sort(unique(shot_zones$player_name))

ui <- fluidPage(
  titlePanel("Lakers Evolution Dashboard"),
  sidebarLayout(
    sidebarPanel(
      width = 3,
      selectInput(
        inputId  = "player",
        label    = "Choose a player",
        choices  = player_choices,
        selected = "Rui Hachimura"
      ),
      p("Hover over a bar to see the exact share and number of attempts.")
    ),
    mainPanel(
      width = 9,
      plotlyOutput("zone_bars", height = "500px")
    )
  )
)

server <- function(input, output, session) {
  output$zone_bars <- renderPlotly({
    plot_data <- shot_zones |>
      filter(player_name == input$player)

    gg <- ggplot(plot_data,
                 aes(x = shot_zone_basic, y = share, fill = era,
                     text = paste0(
                       era, "<br>",
                       shot_zone_basic, "<br>",
                       "Attempts: ", n, "<br>",
                       "Share: ", round(share * 100, 1), "%"
                     ))) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      scale_fill_manual(values = lakers_colours) +
      labs(title = paste0(input$player, ": shot distribution by zone"),
           x = NULL, y = "Share of attempts", fill = NULL) +
      theme_minimal(base_size = 12) +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))

    ggplotly(gg, tooltip = "text")
  })
}

shinyApp(ui = ui, server = server)
```

Run the app. The chart is now a Plotly chart *inside* a Shiny app — the dropdown still changes the player, and every bar is also hoverable.

That is the full shape of a first Shiny app:

``` text
input --> server --> ggplot (+ ggplotly) --> plotlyOutput
```

::: callout-tip
Shiny has an official "Gallery" where you can get inspirations to improve your outputs! \[[link](https://shiny.posit.co/r/gallery/)\]
:::

## What to Notice

Three ideas underneath the whole workbook are worth flagging:

1.  **Derive the variable you want to compare before plotting.** `era` is the whole reason this is an *evolution* dashboard. Everything else is just bars on top of that one derived column.
2.  **Build up in stages.** Static → Plotly → Shiny v1 → v2 → v3 → v4. Each stage added one idea, and each stage was testable on its own.
3.  **Reactivity is simple at its core.** An output that reads `input$something` re-runs whenever that input changes. That one sentence captures almost all of introductory Shiny.

## Independent Practice

Try these small extensions:

1.  change the `selected` player to someone else and re-run the app;
2.  swap `share` for the **raw attempt count** (`n`) on the y-axis, and compare what the two versions emphasise;
3.  compute **FG% per zone** instead of share — rebuild the pre-aggregation step from `lakers_shots` using `summarise(fg_pct = mean(shot_made_flag))`, save that as a new RDS and plot it;
4.  add a `checkboxInput` in the UI called `"made_only"` — you will need to rebuild the pre-aggregation with a `made` flag so the app can switch between "all shots" and "made only";
5.  change the colour palette and the title to match a different team.

## Common Issues

| Problem | Solution |
|------------------------------------|------------------------------------|
| `readRDS()` cannot find the file | Check that the Canvas file is saved in `data/cleaned/` and that the pre-aggregation step wrote `app/shot_zones.rds` |
| App launches but the chart is blank | Check the player name matches one in the data — `input$player` is case-sensitive |
| Bars appear in the wrong order | Re-run the pre-aggregation step — `shot_zone_basic` needs to be a factor with `zone_order` as its levels in `lakers_shots` before you aggregate |
| Percentages print as `0.25` | Add `scale_y_continuous(labels = scales::percent_format())` |
| Colours do not match the Lakers | Check that `lakers_colours` is defined and passed via `scale_fill_manual()` |
| Plotly chart does not appear in Shiny | Confirm `plotlyOutput()` in the UI pairs with `renderPlotly()` in the server |
| The dropdown is empty | Check that `player_choices` is built from the data before the `ui` block runs |

## Before You Go

Great work! ✨

Today you:

-   started from a cleaned `.rds` file of Lakers shots across four regular seasons;
-   split those seasons into an AD Era and a Luka Era around the 1 February 2025 trade;
-   built a static bar chart comparing a player's shot distribution across zones;
-   converted the chart into a Plotly chart with hover text;
-   **pre-aggregated** the data into a 60-row `shot_zones.rds` so the Shiny app stays lightweight;
-   built a first Shiny app in four small stages — minimal, with a selector, with a layout, and finally with Plotly inside Shiny.

Next week we will turn this app into a more polished dashboard with a richer layout, additional charts and a short automatic commentary panel.