---
title: "4. Interactive Dashboards with Shinylive II"
---

Last week's Shiny app was functional — a sidebar, a dropdown and a single Plotly bar chart — but it read like *one chart with controls*, not a dashboard. This week we stay with the same Lakers Evolution story and close that **design gap**. Same data, same before/after question around the 1 February 2025 Luka trade, new layer of polish.

We will keep the bar chart from last week and add three more angles on the same story:

-   an **efficiency companion chart** — the same zones on the x-axis, but now the bars show eFG% instead of attempts — paired with the volume chart inside a tabbed card;
-   a **second card** with a single interactive plotly tree map — the original `action_type` strings nested inside families (Dunk, Layup, Fadeaway, Jump Shot) and eras, tile area = share of that era's attempts, tile colour = eFG%, detail on hover only;
-   three efficiency **metrics** — `eFG%`, `xeFG%` and the `eFG% − xeFG%` differential — shown as headline value boxes that compare the Luka Era to the AD Era at a glance.

Each visual is **sketched as a static figure first**, then picked up and dropped into the dashboard. The real teaching goal is the layout toolkit that `shiny` and `bslib` give us — `page_sidebar()`, `card()`, `navset_card_tab()`, `value_box()` and `layout_columns()`. The workflow for today is:

``` text
clean data -> save final product to app/ -> sketch each chart statically
                                         -> learn layout patterns
                                         -> assemble the dashboard
```

Only one file ever leaves this workbook — the single cleaned `.rds` the app reads.

## Reading List

-   Posit Official Layouts Tutorial: \[[link](https://shiny.posit.co/r/layouts/)\]
-   Wickham, H. 2021. *Mastering Shiny*. \[[link](https://mastering-shiny.org/)\]
-   The Shiny Cheat Sheet: \[[link](https://shiny.posit.co/r/articles/start/cheatsheet/)\]
-   `bslib` dashboards reference: \[[link](https://rstudio.github.io/bslib/articles/dashboards/index.html)\]

## Setup

We keep last week's stack and add `bslib` for the layout components, `tidyr` for the wide-format era table the value boxes read from, and `treemapify` for the static tree-map sketch. The interactive tree map is built with `plotly` directly — no extra library needed.

```{r}
#| message: false
#| warning: false
#| filename: "R code"

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
library(shiny)
library(bslib)
library(treemapify)
```

## Three New Metrics

Before we open the app, we add three numbers to the story. Each answers a different question — together, they let us tell shot-making apart from shot-selection.

| Metric           | Question it answers                                |
|------------------|----------------------------------------------------|
| **eFG%**         | Was the shooting actually effective?               |
| **xeFG%**        | Were the shots good ones to take? *(shot quality)* |
| **eFG% − xeFG%** | Did the player finish them? *(shot-making)*        |

**eFG%** is standard field-goal percentage, but a made 3-pointer counts as 1.5 makes — so a 5/10 shooter who took all 3s lands at 75%, not 50%.

**xeFG%** is the eFG% an *average* shooter would produce given this player's exact shot mix. High xeFG% = easy menu (dunks, layups, open 3s). Low xeFG% = hard menu (contested pull-ups, fadeaways).

**The differential** (`eFG% − xeFG%`) is shot-making stripped of shot-selection. Positive = finishing better than the shot difficulty predicts; negative = worse. Report it in **percentage points (pp)**, not `%`.

::: {.callout-note collapse="true"}
## How each metric is computed

**eFG% — Effective Field Goal %**

A made 3 is worth 50% more than a made 2, so eFG% credits it accordingly:

``` text
eFG% = (FGM + 0.5 × 3PM) / FGA
```

In our data we use a compact equivalent that uses `shot_value` (2 or 3) and `shot_made_flag` (0 or 1):

``` r
efg = sum(shot_made_flag * shot_value) / (2 * n())
```

Worked example — a player takes 10 shots (4 RA at 3/4, 4 Mid-Range at 2/4, 2 Above-the-Break 3s at 1/2):

``` text
numerator   = 3×2 + 2×2 + 1×3 = 13
denominator = 2 × 10           = 20
eFG%        = 13 / 20          = 65%
```

A perfect 2-point shooter tops out at 100%; a perfect 3-point shooter at 150%.

**xeFG% — Expected eFG%**

Two things set a shot's difficulty: `shot_zone_basic` (where) and `action_type` (how). We compute xeFG% in two steps.

*Step 1 — baseline per (zone × action_type).* Over every shot in the dataset:

``` r
shot_baseline <- lakers_shots |>
  group_by(shot_zone_basic, action_type) |>
  summarise(xefg_baseline = sum(shot_made_flag * shot_value) / (2 * n()))
```

One number per combination, e.g. Restricted Area + Dunk Shot ≈ 0.93; Mid-Range + Fadeaway ≈ 0.33.

*Step 2 — join and average per player.* Attach each shot's baseline back onto the shot file, then average across the player's attempts:

``` r
player_xefg = mean(xefg_baseline)
```

Because every shot carries its own baseline, the mean is a **shot-weighted average** — shots the player takes often dominate the number.

**Differential — eFG% − xeFG%**

A simple subtraction. For the worked example with xeFG% = 50.4%:

``` text
differential = 65.0% − 50.4% = +14.6 pp
```
:::

::: {.callout-tip collapse="true"}
## Reading the three metrics together

What each extreme *actually tells you*:

-   **eFG% — high**: the player is scoring effectively. Makes are landing, and 3-pointers (if any) are getting the extra credit they deserve.
-   **eFG% — low**: not enough shots are falling. Could be hard shots, could be poor shot-making — eFG% alone cannot tell you which.
-   **xeFG% — high**: the player is taking *easy* shots on average — think dunks, layups, open catch-and-shoot 3s, Restricted Area finishes. Their shot *selection* looks comfortable.
-   **xeFG% — low**: the player is taking *hard* shots on average — contested pull-ups, long 2s, step-back fadeaways. Tough menu.
-   **Differential — positive (eFG% \> xeFG%)**: they are making **more** than the shot difficulty predicts. Either they are a genuinely great finisher/shooter, or they are on a hot stretch.
-   **Differential — negative (eFG% \< xeFG%)**: they are making **less** than the shot difficulty predicts. Cold stretch, poor finishing, or defensive pressure the baseline does not see.
-   **Differential — near zero**: they are making roughly what the shot difficulty predicts. Their efficiency is explained entirely by *what* they attempt, not by unusual shot-making.

And the pairs worth watching:

-   **High eFG% + high xeFG%** → efficient but on easy shots. Read it as *"role player doing their job"*, not *"star shooter"*.
-   **High eFG% + low xeFG%** → efficient *and* on hard shots. That is a star-shooter signal — the differential will be large and positive.
-   **Low eFG% + high xeFG%** → easy shots but still missing them. A red flag — the baseline says those shots should be going in.
-   **Low eFG% + low xeFG%** → hard shots *and* not finishing them. Often a shot-selection problem disguised as a shot-making one.
:::

## Prepare the Cleaned App Dataset

The data-prep step is the same habit as last week: do the heavy lifting here, and ship only what the app needs. The *only* file we will write into `app/` is the finished product.

```{r}
#| filename: "R code"

luka_trade_date <- as.Date("2025-02-01")

zone_order <- c(
  "Restricted Area",
  "In The Paint (Non-RA)",
  "Mid-Range",
  "Left Corner 3",
  "Right Corner 3",
  "Above the Break 3"
)

lakers_shots <- readRDS("data/cleaned/lakers_shots.rds") |>
  mutate(
    game_date = as.Date(game_date),
    era = factor(
      if_else(game_date < luka_trade_date, "AD Era", "Luka Era"),
      levels = c("AD Era", "Luka Era")
    )
  ) |>
  filter(shot_zone_basic %in% zone_order) |>
  mutate(shot_zone_basic = factor(shot_zone_basic, levels = zone_order))

shot_baseline <- lakers_shots |>
  group_by(shot_zone_basic, action_type) |>
  summarise(xefg_baseline = sum(shot_made_flag * shot_value) / (2 * n()),
            .groups = "drop")

lakers_shots <- lakers_shots |>
  left_join(shot_baseline, by = c("shot_zone_basic", "action_type"))
```

`shot_baseline` is the xeFG% an average shooter would produce for each (zone × action_type) combination, computed once from every shot in the dataset. Joining it back onto `lakers_shots` adds an `xefg_baseline` column to every row, so the app never has to recompute the baseline.

Save the finished table into `app/`.

```{r}
#| eval: false
#| filename: "R code"

dir.create("app", showWarnings = FALSE)
saveRDS(lakers_shots, "app/lakers_shots.rds")
```

```{r}
#| include: false

dir.create("app", showWarnings = FALSE)
saveRDS(lakers_shots, "app/lakers_shots.rds")
```

That is the only file the app will read this week. Your project should now look like this:

``` text
app/
  app.R
  lakers_shots.rds          <- only this file moves to the app folder
data/
  cleaned/
    lakers_shots.rds        <- the original cleaned archive stays here
```

## Sketch the Static Visuals

Every chart the dashboard shows is going to be a chart we have already built and inspected on its own. Four sketches, four charts — two zone views, two action-type views, plus the per-era metric table.

### Sketch 1: Shot-Zone Bar Chart

Reuse the chart from last week. `share` is the proportion of Rui's attempts that came from each zone within each era.

```{r}
#| label: fig-rui-static-bars
#| fig-cap: "Share of attempts per zone, AD Era vs Luka Era."
#| filename: "R code"

lakers_colours <- c("AD Era" = "#FDB927", "Luka Era" = "#552583")

rui_by_zone <- lakers_shots |>
  filter(player_name == "Rui Hachimura") |>
  count(era, shot_zone_basic) |>
  group_by(era) |>
  mutate(share = n / sum(n)) |>
  ungroup()

ggplot(rui_by_zone, aes(x = shot_zone_basic, y = share, fill = era)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = lakers_colours) +
  labs(x = NULL, y = "Share of attempts", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

### Sketch 2: Zone Efficiency (Volume's Companion)

Sketch 1 answers *where did the shots come from?*. The obvious follow-up is *how well did they go in?*. Notice the code below is almost identical to Sketch 1 — same zones on the x-axis, same era split, same Lakers colours — but the numerator changes from **share of attempts** to **eFG% per zone**. That is the teaching point: once you have built one chart, a close sibling is often ten lines away.

```{r}
#| label: fig-rui-static-zone-eff
#| fig-cap: "eFG% per zone, AD Era vs Luka Era. Same structure as Sketch 1, different metric."
#| filename: "R code"

rui_zone_eff <- lakers_shots |>
  filter(player_name == "Rui Hachimura") |>
  group_by(era, shot_zone_basic) |>
  summarise(efg      = sum(shot_made_flag * shot_value) / (2 * n()),
            attempts = n(),
            .groups  = "drop")

ggplot(rui_zone_eff, aes(x = shot_zone_basic, y = efg, fill = era)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = lakers_colours) +
  labs(x = NULL, y = "eFG%", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

Paired together, the two charts tell the classic *volume + efficiency* story:

-   **Volume chart** — *which zones did Rui shoot from, and did the mix change?*
-   **Efficiency chart** — *did he make more of them from those places?*

That pairing is what makes Pattern B (a tabbed card) the natural home for these two later on. Keep the `attempts` column on `rui_zone_eff`; we will use it for the Plotly tooltip inside the app.

### Sketch 3: Action Families as Tree Maps

`shot_zone_basic` told us *where*; `action_type` tells us *how*. The catch: NBA shot data carries 30+ distinct `action_type` strings — "Driving Reverse Layup", "Pullup Jump Shot", "Alley Oop Dunk", "Running Fadeaway Shot"... A bar chart across that many categories is unreadable. A **tree map** copes because unimportant categories shrink to thin slivers automatically; the reader's eye lands on the big tiles first.

We will keep the *original* action_type strings as the tile labels — they carry information — and only use the **family** as the subgroup hierarchy above them. Four families, plus an Other bucket for the stragglers:

**Step 1 — the family function.** If a string matches more than one keyword (e.g. "Fadeaway Jump Shot" contains both Fadeaway and Jump), precedence is Dunk → Layup → Fadeaway → Jump. The first match wins.

```{r}
#| filename: "R code"

action_family_of <- function(x) {
  dplyr::case_when(
    grepl("Dunk",     x, ignore.case = TRUE) ~ "Dunk",
    grepl("Layup",    x, ignore.case = TRUE) ~ "Layup",
    grepl("Fadeaway", x, ignore.case = TRUE) ~ "Fadeaway",
    grepl("Jump",     x, ignore.case = TRUE) ~ "Jump Shot",
    TRUE                                     ~ "Other"
  )
}
```

**Step 2 — the per-era, per-action table.** One row per (era × action_family × action_type) combination.

```{r}
#| filename: "R code"

rui_family <- lakers_shots |>
  filter(player_name == "Rui Hachimura") |>
  mutate(action_family = action_family_of(action_type)) |>
  group_by(era, action_family, action_type) |>
  summarise(attempts = n(),
            efg      = sum(shot_made_flag * shot_value) / (2 * n()),
            .groups  = "drop") |>
  group_by(era) |>
  mutate(proportion = attempts / sum(attempts)) |>
  ungroup()

rui_family
```

Each row carries `attempts`, `proportion` and `efg`. The tree map sizes tiles by `proportion` (share *within* each era) so the two era sub-trees stay comparable even when one has fewer total shots, and colours tiles by `efg`.

**Step 3 — the static sketch.** `treemapify` gives ggplot a `geom_treemap()` we can drop straight onto `rui_family`. Facet per era so the two halves stay visually separated, and keep the scale-fill gradient on `efg`.

```{r}
#| label: fig-rui-static-treemap
#| fig-cap: "Rui Hachimura's action-type tree map, AD Era vs Luka Era. Tile size = share within era, fill = eFG%."
#| filename: "R code"

ggplot(rui_family,
       aes(area    = proportion,
           fill    = efg,
           label   = action_type,
           subgroup = action_family)) +
  geom_treemap() +
  geom_treemap_subgroup_border(colour = "white", size = 3) +
  geom_treemap_text(colour = "white", place = "centre",
                    grow = FALSE, reflow = TRUE, min.size = 6) +
  scale_fill_viridis_c(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 1), name = "eFG%") +
  facet_wrap(~ era) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))
```

The static sketch already tells the story: which families dominate each era, and which individual action_types inside them are getting finished. The interactive plotly version in the app will add hover tooltips and drill-down, but the **shape** of the chart is the one you see here.

::: {.callout-note collapse="true"}
## Why a tree map (and why build it interactively)?

`action_type` in NBA shot data has a long tail: 30+ distinct strings, most attempted only a handful of times per player. Two common reflexes both fail:

-   a **bar chart** with 30 dodged categories crashes the x-axis labels and shrinks the bars we most want to read;
-   a **top-N filter** hides exactly the specialised shots that separate one shooter from another (a Step-Back 3, a Turnaround Fadeaway).

A tree map sidesteps both: tile **area** encodes share of attempts — common shots dominate, rare ones shrink to slivers — and a **subgroup hierarchy** (family → action_type) lets the eye land on a family first. The static sketch above nails the layout; the interactive plotly version in the app adds hover tooltips with attempts, eFG% and the change vs the other era.
:::

`rui_family` is the only structure the app needs; the plotly code lives with the rest of the dashboard in the next section.

### Sketch 4: The Three Headline Numbers, Per Era

The three metrics become *a story* when we compute them **once per era** and compare. That is the shape the value boxes will end up showing: the Luka Era number as the headline, with a short "▲ +4.1 pp vs AD Era" line underneath. Build the table first.

```{r}
#| filename: "R code"

rui_era_metrics <- lakers_shots |>
  filter(player_name == "Rui Hachimura") |>
  group_by(era) |>
  summarise(attempts = n(),
            efg      = sum(shot_made_flag * shot_value) / (2 * n()),
            xefg     = mean(xefg_baseline),
            diff     = efg - xefg,
            .groups  = "drop")

rui_era_metrics
```

Three columns matter downstream:

-   `efg` — the number the first value box will show.
-   `xefg` — the second value box; reads "shot quality".
-   `diff` — the third value box; reads "shot-making above/below expectation".

The **delta** shown under each value box is just the Luka row minus the AD row, expressed in percentage points. We will compute it in the app with a short helper. Having the era table in front of you now makes it obvious what those helpers need to produce.

Four charts, one pattern in common: each was built, inspected and coloured in the workbook before touching Shiny. If any of them looked wrong here, it would be obvious *here* — not three cards deep in an app.

## Build the Dashboard, Step by Step

We do not build the dashboard in one go. We start from last week's app and add one piece at a time. After each step the app still runs — that way, if something breaks, the cause is the change you *just* made, not six things bundled together.

Each step below shows only the **UI** change (the server pieces each step needs appear in the full listing at the end). Drop each step's code into your own `app/app.R` and run it before moving on.

### Step 0 — Last Week's Starting Point

Last week you had a `fluidPage()` with a sidebar and one Plotly bar chart. Strip it to its bones:

```{r}
#| eval: false
#| filename: "Step 0"

ui <- fluidPage(
  titlePanel("Lakers Evolution Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput("player", "Player",
                  choices = player_choices, selected = "Rui Hachimura")
    ),
    mainPanel(
      plotlyOutput("zone_bars", height = "440px")
    )
  )
)
```

### Step 1 — Swap to `page_sidebar()` + `card()`

`bslib` replaces `fluidPage()` + `sidebarLayout()` with one call that already looks like a dashboard. Frame the chart with `card()` so it picks up padding, a header and a border for free.

```{r}
#| eval: false
#| filename: "Step 1"

ui <- page_sidebar(
  title   = "Lakers Evolution Dashboard",
  sidebar = sidebar(
    width = 260,
    selectInput("player", "Player",
                choices = player_choices, selected = "Rui Hachimura")
  ),
  card(
    card_header("Shot Distribution by Zone"),
    plotlyOutput("zone_bars", height = "440px")
  )
)
```

Same one chart, but the page now has the dashboard look.

### Step 2 — Turn the Card Into a Tabbed Card

Sketches 1 and 2 are two views of the same question. `navset_card_tab()` lets the reader flip between them inside a single card.

```{r}
#| eval: false
#| filename: "Step 2"

navset_card_tab(
  title = "Shot Selection by Zone",
  full_screen = TRUE,
  nav_panel("Volume",     plotlyOutput("zone_bars", height = "360px")),
  nav_panel("Efficiency", plotlyOutput("zone_eff",  height = "360px"))
)
```

`full_screen = TRUE` adds an expand button. Both outputs still react to the same `input$player`; the server has no idea the tabs exist.

### Step 3 — Add a Headline Row of Value Boxes

`value_box()` is a coloured card built around one big number. `layout_columns()` places three in a row. Anything after `value = ...` renders below the number — which is where we slip in the delta line.

```{r}
#| eval: false
#| filename: "Step 3"

layout_columns(
  col_widths = c(4, 4, 4),
  value_box(title    = "eFG%",
            value    = textOutput("efg_value"),
            textOutput("efg_delta"),
            showcase = bsicons::bs_icon("bullseye"),
            theme    = "primary"),
  value_box(title    = "xeFG% (shot quality)",
            value    = textOutput("xefg_value"),
            textOutput("xefg_delta"),
            showcase = bsicons::bs_icon("map"),
            theme    = "secondary"),
  value_box(title    = "eFG% − xeFG% (shot-making)",
            value    = textOutput("diff_value"),
            textOutput("diff_delta"),
            showcase = bsicons::bs_icon("graph-up"),
            theme    = "success")
)
```

Put this row *above* the zone card inside `page_sidebar()`. The server side needs one reactive (the per-era metric table) and six text outputs (three values + three deltas).

### Step 4 — Add the Action-Family Tree Map Side by Side

Wrap the zone tabbed card **and** a new single-chart card in `layout_columns(col_widths = c(6, 6))` so they share a row.

```{r}
#| eval: false
#| filename: "Step 4"

layout_columns(
  col_widths = c(6, 6),
  navset_card_tab(
    title = "Shot Selection by Zone",
    full_screen = TRUE,
    nav_panel("Volume",     plotlyOutput("zone_bars", height = "320px")),
    nav_panel("Efficiency", plotlyOutput("zone_eff",  height = "320px"))
  ),
  card(
    card_header("Action Families — Efficiency"),
    full_screen = TRUE,
    tags$small(style = "color: #666; padding: 0 0.75rem;",
               "Tile size = share of attempts within each era, ",
               "tile colour = eFG%. Hover for details."),
    plotlyOutput("type_tree_efficiency", height = "320px")
  )
)
```

When the column widths add up to 12, Bootstrap places them side by side.

### Step 5 — Theme and Polish

A single `bs_theme()` propagates the Lakers palette to every component. Narrow the sidebar so the two cards get room, and shrink the value-box typography so the headline row reads as a summary rather than the hero.

```{r}
#| eval: false
#| filename: "Step 5"

lakers_theme <- bs_theme(
  version   = 5,
  primary   = "#552583",
  secondary = "#FDB927",
  success   = "#2E7D32",
  base_font = font_google("Inter")
)

vb_title <- function(x) tags$span(style = "font-size: 0.8rem;",  x)
vb_value <- function(x) tags$span(style = "font-size: 1.3rem; font-weight: 600;", x)
vb_delta <- function(x) tags$span(style = "font-size: 0.75rem;", x)
vb_icon  <- function(name) tags$span(style = "font-size: 0.4rem;",
                                     bsicons::bs_icon(name))

# in page_sidebar: theme = lakers_theme, sidebar = sidebar(width = 200, ...)
# in value_box:    wrap title/value/delta/showcase with the helpers above
```

Set the theme once, pass it via `page_sidebar(theme = lakers_theme)`, and every `value_box(theme = "primary")`, active tab and button matches — no `tags$style()` hacks required.

### The Finished `app.R`

Every step above collapses into one file. Copy this into `app/app.R`:

```{r}
#| eval: false
#| filename: "app/app.R"

library(shiny)
library(plotly)
library(dplyr)
library(tidyr)
library(ggplot2)
library(bslib)

lakers_colours  <- c("AD Era" = "#FDB927", "Luka Era" = "#552583")
luka_trade_date <- as.Date("2025-02-01")

lakers_shots   <- readRDS("lakers_shots.rds")
player_choices <- sort(unique(lakers_shots$player_name))

action_family_of <- function(x) {
  case_when(
    grepl("Dunk",     x, ignore.case = TRUE) ~ "Dunk",
    grepl("Layup",    x, ignore.case = TRUE) ~ "Layup",
    grepl("Fadeaway", x, ignore.case = TRUE) ~ "Fadeaway",
    grepl("Jump",     x, ignore.case = TRUE) ~ "Jump Shot",
    TRUE                                     ~ "Other"
  )
}

lakers_theme <- bs_theme(
  version   = 5,
  primary   = "#552583",
  secondary = "#FDB927",
  success   = "#2E7D32",
  base_font = font_google("Inter")
)

vb_title <- function(x) tags$span(style = "font-size: 0.8rem;",  x)
vb_value <- function(x) tags$span(style = "font-size: 1.3rem; font-weight: 600;", x)
vb_delta <- function(x) tags$span(style = "font-size: 0.75rem;", x)
vb_icon  <- function(name) tags$span(style = "font-size: 0.4rem;",
                                     bsicons::bs_icon(name))

ui <- page_sidebar(
  title   = "Lakers Evolution Dashboard",
  theme   = lakers_theme,
  sidebar = sidebar(
    width = 200,
    selectInput("player", "Player",
                choices  = player_choices,
                selected = "Rui Hachimura"),
    p("Value boxes compare the Luka Era to the AD Era.
       The left card splits shots by court zone.
       The right card is an interactive tree map — hover any tile
       for attempts, share, eFG% and the change vs the other era.")
  ),

  # --- Pattern C: headline numbers with era deltas (slim) ---
  layout_columns(
    col_widths = c(4, 4, 4),
    value_box(title    = vb_title("eFG% (efficiency)"),
              value    = vb_value(textOutput("efg_value", inline = TRUE)),
              vb_delta(textOutput("efg_delta", inline = TRUE)),
              showcase = vb_icon("bullseye"),
              theme    = "primary"),
    value_box(title    = vb_title("xeFG% (shot quality)"),
              value    = vb_value(textOutput("xefg_value", inline = TRUE)),
              vb_delta(textOutput("xefg_delta", inline = TRUE)),
              showcase = vb_icon("map"),
              theme    = "secondary"),
    value_box(title    = vb_title("eFG% − xeFG% (shot-making)"),
              value    = vb_value(textOutput("diff_value", inline = TRUE)),
              vb_delta(textOutput("diff_delta", inline = TRUE)),
              showcase = vb_icon("graph-up"),
              theme    = "success")
  ),

  # --- Pattern D: two cards side by side ---
  layout_columns(
    col_widths = c(6, 6),

    # Pattern B: zone — Volume + Efficiency in a tabbed card
    navset_card_tab(
      title = "Shot Selection by Zone",
      full_screen = TRUE,
      nav_panel("Volume",     plotlyOutput("zone_bars", height = "320px")),
      nav_panel("Efficiency", plotlyOutput("zone_eff",  height = "320px"))
    ),

    # Action families — single-chart card, tree map of original action_types
    card(
      card_header("Action Families — Efficiency"),
      full_screen = TRUE,
      tags$small(style = "color: #666; padding: 0 0.75rem;",
                 "Tile size = share of attempts within each era, ",
                 "tile colour = eFG%. Hover for details."),
      plotlyOutput("type_tree_efficiency", height = "320px")
    )
  )
)

server <- function(input, output, session) {

  player_shots <- reactive({
    lakers_shots |> filter(player_name == input$player)
  })

  # --- Era metrics (Sketch 4) ---
  era_metrics <- reactive({
    player_shots() |>
      group_by(era) |>
      summarise(efg  = sum(shot_made_flag * shot_value) / (2 * n()),
                xefg = mean(xefg_baseline),
                diff = efg - xefg,
                .groups = "drop") |>
      pivot_wider(names_from = era,
                  values_from = c(efg, xefg, diff))
  })

  fmt_pct   <- function(x) paste0(round(x * 100, 1), "%")
  fmt_pp    <- function(x) paste0(if (x >= 0) "+" else "",
                                  round(x * 100, 1), " pp")
  fmt_delta <- function(luka, ad) {
    d <- luka - ad
    arrow <- if (d > 0) "▲" else if (d < 0) "▼" else "—"
    paste0(arrow, " ", fmt_pp(d), " vs AD (", fmt_pct(ad), ")")
  }

  # Headline values = Luka Era
  output$efg_value  <- renderText({ fmt_pct(era_metrics()$`efg_Luka Era`)  })
  output$xefg_value <- renderText({ fmt_pct(era_metrics()$`xefg_Luka Era`) })
  output$diff_value <- renderText({ fmt_pp(era_metrics()$`diff_Luka Era`)  })

  # Delta lines = Luka − AD
  output$efg_delta  <- renderText({
    m <- era_metrics(); fmt_delta(m$`efg_Luka Era`,  m$`efg_AD Era`)
  })
  output$xefg_delta <- renderText({
    m <- era_metrics(); fmt_delta(m$`xefg_Luka Era`, m$`xefg_AD Era`)
  })
  output$diff_delta <- renderText({
    m <- era_metrics(); fmt_delta(m$`diff_Luka Era`, m$`diff_AD Era`)
  })

  # --- Volume: share of attempts per zone (Sketch 1) ---
  output$zone_bars <- renderPlotly({
    plot_data <- player_shots() |>
      count(era, shot_zone_basic) |>
      group_by(era) |>
      mutate(share = n / sum(n)) |>
      ungroup()

    gg <- ggplot(plot_data,
                 aes(x = shot_zone_basic, y = share, fill = era,
                     text = paste0(era, "<br>", shot_zone_basic, "<br>",
                                   "Attempts: ", n, "<br>",
                                   "Share: ", round(share * 100, 1), "%"))) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      scale_fill_manual(values = lakers_colours) +
      labs(x = NULL, y = "Share of attempts", fill = NULL) +
      theme_minimal(base_size = 11) +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))

    ggplotly(gg, tooltip = "text")
  })

  # --- Efficiency: eFG% per zone (Sketch 2) ---
  output$zone_eff <- renderPlotly({
    plot_data <- player_shots() |>
      group_by(era, shot_zone_basic) |>
      summarise(efg      = sum(shot_made_flag * shot_value) / (2 * n()),
                attempts = n(),
                .groups  = "drop")

    gg <- ggplot(plot_data,
                 aes(x = shot_zone_basic, y = efg, fill = era,
                     text = paste0(era, "<br>", shot_zone_basic, "<br>",
                                   "Attempts: ", attempts, "<br>",
                                   "eFG%: ", round(efg * 100, 1), "%"))) +
      geom_col(position = position_dodge(width = 0.8), width = 0.7) +
      scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
      scale_fill_manual(values = lakers_colours) +
      labs(x = NULL, y = "eFG%", fill = NULL) +
      theme_minimal(base_size = 11) +
      theme(legend.position = "top",
            axis.text.x = element_text(angle = 20, hjust = 1))

    ggplotly(gg, tooltip = "text")
  })

  # --- Action-family long table, one row per (era × family × action_type).
  #     proportion is the share of attempts *within each era*, so the two
  #     era sub-trees stay comparable even when one has fewer total shots. ---
  family_data <- reactive({
    player_shots() |>
      mutate(action_family = action_family_of(action_type)) |>
      group_by(era, action_family, action_type) |>
      summarise(attempts = n(),
                efg      = sum(shot_made_flag * shot_value) / (2 * n()),
                .groups  = "drop") |>
      group_by(era) |>
      mutate(proportion = attempts / sum(attempts)) |>
      ungroup()
  })

  # --- Build one treemap-node table per era: families at the root,
  #     original action_types as leaves. The era label is handled as a
  #     subplot title (not as a tile) so the drill-down starts at family. ---
  era_tree_nodes <- reactive({
    d <- family_data()

    wide <- d |>
      mutate(era_key = if_else(era == "AD Era", "ad", "luka")) |>
      select(-era) |>
      pivot_wider(id_cols     = c(action_family, action_type),
                  names_from  = era_key,
                  values_from = c(attempts, proportion, efg))

    enriched <- d |>
      left_join(wide, by = c("action_family", "action_type")) |>
      mutate(
        other_era        = if_else(era == "AD Era", "Luka Era", "AD Era"),
        proportion_other = if_else(era == "AD Era",
                                   coalesce(proportion_luka, 0),
                                   coalesce(proportion_ad,   0)),
        efg_other        = if_else(era == "AD Era", efg_luka, efg_ad),
        proportion_delta = proportion - proportion_other,
        efg_delta        = efg - efg_other
      )

    sign_pp <- function(x) paste0(if_else(x >= 0, "+", ""),
                                  round(x * 100, 1), " pp")

    build_one <- function(era_name) {
      e <- enriched |> filter(era == era_name)
      if (nrow(e) == 0) return(NULL)

      leaves <- e |>
        transmute(
          id      = paste(action_family, action_type, sep = "/"),
          label   = action_type,
          parent  = action_family,
          value   = proportion,
          efg_val = efg,
          hover   = paste0(
            "<b>", action_type, "</b><br>",
            "Attempts: ", attempts, "<br>",
            "Share: ", round(proportion * 100, 1), "%",
            "  (", sign_pp(proportion_delta), " vs other era)<br>",
            "eFG%: ", round(efg * 100, 1), "%",
            if_else(!is.na(efg_delta),
                    paste0("  (", sign_pp(efg_delta), " vs other era)"),
                    "")
          )
        )

      families <- e |>
        group_by(action_family) |>
        summarise(attempts_total   = sum(attempts),
                  proportion_total = sum(proportion),
                  efg_val          = weighted.mean(efg, attempts,
                                                   na.rm = TRUE),
                  .groups          = "drop") |>
        transmute(
          id      = action_family,
          label   = action_family,
          parent  = "",
          value   = proportion_total,
          efg_val = efg_val,
          hover   = paste0("<b>", action_family, "</b><br>",
                           "Attempts: ", attempts_total, "<br>",
                           "Share: ", round(proportion_total * 100, 1), "%<br>",
                           "eFG%: ", round(efg_val * 100, 1), "%")
        )

      bind_rows(families, leaves)
    }

    list(ad = build_one("AD Era"), luka = build_one("Luka Era"))
  })

  # --- Tree map: Efficiency.
  #     Two traces (AD Era, Luka Era) side by side with era as the title;
  #     tile area = share of attempts within that era;
  #     tile colour = eFG% (shared gradient); default labels show on tiles
  #     big enough to hold them, full detail is on hover. ---
  output$type_tree_efficiency <- renderPlotly({
    trees <- era_tree_nodes()
    validate(need(!is.null(trees$ad) || !is.null(trees$luka),
                  "No shots to show for this player."))

    efg_marker <- function(efg_vals) list(
      colors     = efg_vals,
      colorscale = "Viridis",
      cmin       = 0,
      cmax       = 1,
      showscale  = TRUE,
      colorbar   = list(title = list(text = "eFG%"),
                        tickformat = ".0%")
    )

    add_era_trace <- function(p, nodes, column) {
      if (is.null(nodes)) return(p)
      add_trace(p,
        type          = "treemap",
        ids           = nodes$id,
        labels        = nodes$label,
        parents       = nodes$parent,
        values        = nodes$value,
        branchvalues  = "total",
        textinfo      = "label",
        text          = nodes$hover,
        hovertemplate = "%{text}<extra></extra>",
        marker        = efg_marker(nodes$efg_val),
        domain        = list(column = column)
      )
    }

    annotations <- list(
      list(text = "<b>AD Era</b>",
           x = 0.22, y = 1.0, xref = "paper", yref = "paper",
           xanchor = "center", yanchor = "bottom",
           showarrow = FALSE, font = list(size = 13)),
      list(text = "<b>Luka Era</b>",
           x = 0.78, y = 1.0, xref = "paper", yref = "paper",
           xanchor = "center", yanchor = "bottom",
           showarrow = FALSE, font = list(size = 13))
    )

    plot_ly() |>
      add_era_trace(trees$ad,   0) |>
      add_era_trace(trees$luka, 1) |>
      layout(
        grid        = list(columns = 2, rows = 1),
        annotations = annotations,
        margin      = list(t = 60, l = 5, r = 5, b = 5)
      )
  })
}

shinyApp(ui = ui, server = server)
```

Run it.

```{r}
#| eval: false
#| filename: "R Console"

shiny::runApp("app")
```

Each server output maps back to a sketch: `zone_bars` → Sketch 1, `zone_eff` → Sketch 2, `type_tree_efficiency` → Sketch 3 (enriched with a cross-era join so the hover can compare), and the six text outputs (three values + three deltas) all read from the era table in Sketch 4.

## Polish That Pays Off

The last mile of a dashboard is not about the charts — it is the *finishing*. A short checklist to run through before calling it done:

-   **Give every chart a card header or tab title.** `"Volume"` and `"Efficiency"` are doing caption work here — short, specific, and they tell the user what is different between the two views.
-   **Keep the colour story tight.** AD Era gold, Luka Era purple. Use the same mapping in both bar charts, the trend line and — via `bs_theme()` — the value-box accents.
-   **Label value boxes for humans.** `"eFG% − xeFG%  (shot-making)"` is long, but it tells the user what the number means. The delta line below it does the comparison work without a legend.
-   **Set a narrow sidebar.** `sidebar(width = 200)` frees horizontal space for the charts — enough for one dropdown and a short caption, no wider.
-   **Turn on `full_screen = TRUE`** for any card a user might want to stare at.
-   **Match heights thoughtfully.** Sharing `height = "320px"` across the two side-by-side cards keeps the page rhythmic — each card is the same size, and neither one dominates.
-   **Shrink value boxes on data-heavy pages.** A value box defaults to a loud headline size. When the row sits above two tall cards, wrap the title/value/delta in `tags$span(style = "font-size: ...")` to keep the eye on the charts below.
-   **Write a short sidebar caption.** One paragraph telling the user what the page shows saves every future visitor a minute of puzzling.

## What to Notice

Four ideas underneath the whole workbook are worth flagging:

1.  **Static first, then Shiny.** Every chart in the dashboard existed as a static figure before the app was opened. Debugging charts outside Shiny is much easier than debugging inside.
2.  **Ship only the finished data product.** One file, `app/lakers_shots.rds`, with every column the app needs. The raw cleaning code stays in the workbook.
3.  **Think in rows of cards.** A dashboard is a grid of components — value boxes, cards, tabs — each doing one job. Layout is a *vocabulary*, not a one-off choice.
4.  **Consistency tells the story.** Two colours, one font, matching labels across panels. A consistent page feels designed; an inconsistent page feels accidental.

## Common Issues

| Problem | Solution |
|------------------------------------|------------------------------------|
| `page_sidebar()` not found | Install and load `bslib` |
| `bsicons` not found | `install.packages("bsicons")` once, then load it |
| Value boxes show `NaN%` | The selected player has no shots in one era — add `req(all(c("AD Era", "Luka Era") %in% era_metrics()$era))` or guard against missing eras in the helpers |
| Delta says `NA pp` | The pivot-wider column names include a space (`"efg_Luka Era"`) — wrap them in backticks when you index |
| Tree map is empty for a player with no shots | `validate(need(...))` fires on empty data — not a bug, the player genuinely has zero attempts in the current filter |
| Big tiles show a label but small tiles don't | That is the plotly treemap default — labels only render where they fit. Hover any tile for the full `action_type`, attempts, share and eFG% |
| Hover tooltip looks cluttered | Edit the `hover` string in `tree_nodes()` — the lines come from a single `paste0()` so each row is easy to drop or reorder |
| Clicking a tile "zooms in" | That is the plotly treemap default — the top-left corner shows a breadcrumb; click it to zoom out |
| One era sub-tree is empty | The player took no shots in that era — common for mid-season arrivals or departures |
| Tree-map tiles look the same size in both eras | Tile area = share of attempts *within* each era, so a player with 20 AD shots and 200 Luka shots still fills both halves — the *delta* vs the other era is in the hover |
| Volume and Efficiency tabs look identical | Check `y = share` in `zone_bars` and `y = efg` in `zone_eff` — easy to leave both pointing at the same column when copy-pasting |
| Card or value-box colours look generic | Check `bs_theme()` is passed via `page_sidebar(theme = ...)` and that themes are named (`"primary"`, `"secondary"`, `"success"`) |
| Value boxes feel too loud at the top of the page | Wrap the `title`, `value` and delta content in `tags$span(style = "font-size: ...")` — the Polish section of the workbook shows the pattern |
| Sidebar is too wide | Set `sidebar(width = 200)` (or smaller) — narrow sidebars give the side-by-side cards room |
| Tabs inside a card look like plain tabs | Use `navset_card_tab()`, not `tabsetPanel()` — the former is the carded variant |

## Before You Go

You are almost there! ⚙️

Today you:

-   prepared one cleaned app dataset;
-   built zone, efficiency and action-family views;
-   added headline value boxes with AD -\> Luka deltas;
-   used `page_sidebar()`, `card()`, `navset_card_tab()`, `value_box()` and `layout_columns()`;
-   polished the dashboard with a consistent Lakers theme.

Next week we shift from *building the dashboard* to **sharing it** — deployment, GitHub Pages and the final portfolio presentation.